---
title: "R-Ladies Melbourne 4th Anniversary"
author: "Emi Tanaka"
date: "15/11/2020"
output: html_document
---


```{r setup, message = FALSE, warning = FALSE}
library(tidyverse) # for dplyr and ggplot2
```

## Making comparisons by groups?

* Check it out with a plot! 
* Make multiples/facets if there's too many observations per group!
* Easily done with `ggplot2`!

```{r no-shadow}
ggplot(ChickWeight, aes(Time, weight, group = Chick)) +
  geom_line(aes(color = Diet)) +
  facet_wrap(~Diet) +
  theme_bw() +
  labs(y = "Weight") +
  scale_color_viridis_d()
```

* But make it even easier to compare by **adding a shadow of all data for each facet**.
* Only takes one extra line in `ggplot2`. Replace data with the variable you are facetting with dropped and change color to gray like in the second line of code below. 

```{r with-shadow}
ggplot(ChickWeight, aes(Time, weight, group = Chick)) +
  geom_line(data = select(ChickWeight, -Diet), color = "gray") + 
  geom_line(aes(color = Diet)) +
  facet_wrap(~Diet) +
  theme_bw() +
  labs(y = "Weight") +
  scale_color_viridis_d()

```

