---
title: Making a cartogram of the 2020 Victorian COVID-19 outbreak
author: Di Cook 
date: '2020-11-16'
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  message = FALSE, 
  warning = FALSE, 
  cache = TRUE,
  eval = TRUE,
  fig.align = "center"
)
```


```{r libraries, echo=FALSE}
library(tidyverse)
library(lubridate)
library(forcats)
library(ozmaps)
library(sf)
library(cartogram)
library(ggthemes)
library(plotly)
library(sugarbag)
```

In Melbourne we have been in a strict lockdown since July. Each week we get our hopes up that restrictions might be eased, and once again these hopes are dashed by the announcement Sunday Oct 25, keeping the restrictions a little longer because of another outbreak in the northwest of the city. The data we have collected here are the case counts by Victorian local government area (LGA) since the beginning of July. We will examine the spatiotemporal distribution of these counts. 

Working with spatial data is always painful! It almost always requires some **ugly** code. Part of the reason for the difficulty is the use of special data objects, that describe maps. There are several different choices, and some packages and tools use one, and others use another, so not all tools work together. The `sf` package is a recent endeavour that helps enormously, but some tools still use other forms, and when you run into errors this might be the reason - it can be hard to tell. Another reason is that map objects can be very large, which makes sense for accurate mapping, but for data analysis and visualisation, we'd rather have smaller, even if slightly inaccurate, spatial objects. It can be helpful to thin out map data before doing further analysis - you need special tools for this, eg `mapshapr`. We don't need this for the exercises here. Another problem commonly encountered is that there are numerous coordinate systems, and types of projections of the 3D globe into a 2D canvas. We have become accustomed to lat/long but like time its an awkward scale to compute on because a translation from E/W and N/S to positive and negative values is needed. More commonly a Universal Transverse Mercator (UTM) is the standard but its far less intuitive to use.  

The code for all the analysis is provided for you. We recommend that you run the code in steps to see what it is doing, why the mutating and text manipulations are necessary. Talk about the code with each other to help you understand it. 


## Getting the data

### COVID-19 counts

COVID-19 data by LGA is available from https://www.covid19data.com.au/victoria. You should find that some variables are type `chr` because "null" has been used to code entries on some days. This needs fixing, and also missings should be converted to 0, with the rationale being that if the value is missing it most likely means there were no cases.


```{r}
# Read the data
# Replace null with 0, for three LGAs
# Convert to long form to join with polygons
# Make the date variables a proper date
# Set NAs to 0, this is a reasonable assumption
covid <- read_csv("https://raw.githubusercontent.com/numbats/eda/master/data/melb_lga_covid.csv") %>%
  mutate(Buloke = as.numeric(ifelse(Buloke == "null", "0", Buloke))) %>%
  mutate(Hindmarsh = as.numeric(ifelse(Hindmarsh == "null", "0", Hindmarsh))) %>%
   mutate(Towong = as.numeric(ifelse(Towong == "null", "0", Towong))) %>%
  pivot_longer(cols = Alpine:Yarriambiack, names_to="NAME", values_to="cases") %>%
  mutate(Date = ydm(paste0("2020/",Date))) %>%
  mutate(cases=replace_na(cases, 0))

# Case counts are cumulative, so take lags to get daily case counts
covid <- covid %>%
  group_by(NAME) %>%
  mutate(new_cases = cases - dplyr::lag(cases))

# Filter to final day, which is cumulative count
covid_cumul <- covid %>% 
  filter(Date == max(Date)) 
```

### Spatial polygons

Now let's get polygon data of Victorian LGAs from the `ozmaps` package. We need to fix some names of LGAs because there are duplicated LGA names, and there is one mismatch in names from the COVID data and the ozmaps data (Colac Otway). If the COVID data had been provided with a unique LGA code it would have helped in merging with the polygon data.

```{r vic_lga}
# Read the LGA data from ozmaps package. 
# This has LGAs for all of Australia. 
# Need to filter out Victoria LGAs, avoiding LGAs 
# from other states with same name, and make the names
# match covid data names. The regex equation is
# removing () state and LGA type text strings
# Good reference: https://r-spatial.github.io/sf/articles/sf1.html
data("abs_lga")
vic_lga <- abs_lga %>%
  mutate(NAME = ifelse(NAME == "Latrobe (M) (Tas.)", "LatrobeM", NAME)) %>%
  mutate(NAME = ifelse(NAME == "Kingston (DC) (SA)", "KingstonSA", NAME)) %>%
  mutate(NAME = ifelse(NAME == "Bayside (A)", "BaysideA", NAME)) %>% 
  mutate(NAME = str_replace(NAME, " \\(.+\\)", "")) %>%
  mutate(NAME = ifelse(NAME == "Colac-Otway", "Colac Otway", NAME)) 
vic_lga <- st_transform(vic_lga, 3395) 
# 3395 is EPSG CRS, equiv to WGS84 mercator, 
# see https://spatialreference.org/ref/epsg/?page=28
# cartogram() needs this to be set
```

### Choropleth map

A choropleth map is made from filling the colour of polygons. 

```{r covid-choropleth, fig.width=10, fig.height=10, out.width="100%"}
# Join covid data to polygon data, remove LGAs with 
# missing values which should leave just Vic LGAs
vic_lga_covid <- vic_lga %>%
  left_join(covid_cumul, by="NAME") %>%
  filter(!is.na(cases))

# Make choropleth map, with appropriate colour palette
choropleth <- ggplot(vic_lga_covid) + 
  geom_sf(aes(fill = cases, label=NAME), colour="white") + 
  scale_fill_distiller("Cases", palette = "YlOrRd",
                       direction=1) + 
  theme_map() +
  theme(legend.position="bottom")
choropleth
#ggplotly(choropleth) # Interactive map
```

## Population-transformed cartogram

### Get population data

 The file [`VIF2019_Population_Service_Ages_LGA_2036.xlsx`](https://github.com/numbats/eda/blob/master/data/VIF2019_Population_Service_Ages_LGA_2036.xlsx)  has been extracted from the [Vic Gov web site](https://www.planning.vic.gov.au/land-use-and-population-research/victoria-in-future/tab-pages/victoria-in-future-data-tables). It is a complicated `xlsx` file, with the data in sheet 3, and starting 13 rows down. The `readxl` package is handy here to extract the population data needed. The code below has extracted only the data needed.
 
```{r population}
pop <- tibble(NAME = c("Alpine","Ararat","Ballarat","Banyule","Bass Coast","Baw Baw","Bayside","Benalla","Boroondara","Brimbank","Buloke","Campaspe","Cardinia","Casey","Central Goldfields","Colac Otway","Corangamite","Darebin","East Gippsland","Frankston","Gannawarra","Glen Eira","Glenelg","Golden Plains","Greater Bendigo","Greater Dandenong","Greater Geelong","Greater Shepparton","Hepburn","Hindmarsh","Hobsons Bay","Horsham","Hume","Indigo","Kingston","Knox","Latrobe","Loddon","Macedon Ranges","Manningham","Mansfield","Maribyrnong","Maroondah","Melbourne","Melton","Mildura","Mitchell","Moira","Monash","Moonee Valley","Moorabool","Moreland","Mornington Peninsula","Mount Alexander","Moyne","Murrindindi","Nillumbik","Northern Grampians","Port Phillip","Pyrenees","Queenscliffe","South Gippsland","Southern Grampians","Stonnington","Strathbogie","Surf Coast","Swan Hill","Towong","Wangaratta","Warrnambool","Wellington","West Wimmera","Whitehorse","Whittlesea","Wodonga","Wyndham","Yarra","Yarra Ranges","Yarriambiack"), 
pop = c(12578,11746.43,103500.3,127447,33464.85,49296.21,102912,13981.3,177276,204190,6284,37596.09,97572.66,312789,13085.32,21362.81,16241,155126,45598.55,139502,10567.15,148583,19758.61,22016,112270.9,160222,239529.9,65071.32,15526.87,5787.223,93445.04,19884.51,207038,16165.73,158937.6,160353.5,74622.36,7559.041,47479.75,122570.7,8674.158,86942,114799.3,146097.1,141422.1,54658,41794.85,29486,192625,122871.1,32668.76,172289.5,161528,19093.7,16738.47,14052.73,64174,11570.29,108627.2,7315.398,2927.166,29120.95,16122.74,111003,10357.01,30465.01,20895.86,6045.765,28592,34243.1,43531.44,3932.907,169641.6,207058,40100,227008,92898.52,155227.4,6742.772))

vic_lga_covid <- vic_lga_covid %>%
  left_join(pop, by="NAME") 

# Compute additional statistics
vic_lga_covid <- vic_lga_covid %>%
  group_by(NAME) %>%
  mutate(cases_per10k = max(cases/pop*10000, 0)) %>%
  ungroup()
```

### Make the cartogram

```{r vic-cartogram, fig.width=10, fig.height=10, out.width="100%"}
vic_lga_covid_carto <- cartogram_cont(vic_lga_covid, "pop")
# This st_cast() is necessary to get plotly to work
vic_lga_covid_carto <- st_cast(vic_lga_covid_carto, "MULTIPOLYGON") 

cartgram <- ggplot(vic_lga_covid_carto) + 
  geom_sf(aes(fill = cases_per10k, label=NAME), colour="white") + 
  scale_fill_distiller("Cases", palette = "YlOrRd",
                       direction=1) + 
  theme_map() +
  theme(legend.position="bottom") 
cartgram 
#ggplotly(cartgram) # Interactive cartogram
```

## Lastly, a hexagon tile map

The hexagon tile map makes tiled hexagons representing the LGAs. You can read more about it in the documentation for the `sugarbag` package at https://srkobakian.github.io/sugarbag/.

```{r hextile, fig.width=10, fig.height=10, out.width="100%"}
# Spatial coordinates need to be in long/lat
vlc_latlong <- st_transform(vic_lga_covid, crs = "+proj=longlat +datum=WGS84")

# Placement of hexmaps depends on position relative to
# Melbourne central
data(capital_cities)
vic_lga_hexmap <- create_hexmap(
  shp = vlc_latlong,
  sf_id = "NAME",
  focal_points = capital_cities, verbose = TRUE)
# This shows the centroids of the hexagons
# ggplot(vic_lga_hexmap, aes(x=hex_long, y=hex_lat)) +
#  geom_point()

# Hexagons are made with the `fortify_hexagon` function
vic_lga_covid_hexmap <- vic_lga_hexmap %>%
  fortify_hexagon(sf_id = "NAME", hex_size = 0.1869) %>%
  left_join(covid_cumul, by="NAME") %>%
  filter(!is.na(cases)) %>%
  left_join(pop, by="NAME") %>%
  group_by(NAME) %>%
  mutate(cases_per10k = max(cases/pop*10000, 0)) %>%
  ungroup()

hexmap <- ggplot() +
  geom_sf(data=vlc_latlong, 
          fill = "grey90", colour = "white", size=0.1) +
  geom_polygon(data=vic_lga_covid_hexmap, 
               aes(x=long, y=lat, group=hex_id, 
                   fill = cases_per10k, 
                   colour = cases_per10k, 
                   label=NAME), 
               size=0.2) +
  scale_fill_distiller("Cases", palette = "YlOrRd",
                       direction=1) +
  scale_colour_distiller("Cases", palette = "YlOrRd",
                       direction=1) +
  theme_map() +
  theme(legend.position="bottom", aspect.ratio=0.7)
hexmap
# ggplotly(hexmap)
```


