---
title: "R tip"
author: "Anna Quaglieri"
date: "11/11/2020"
output: html_document
---


# Column-wise operations with `mutate`, `across` and `case_when`

I found the method below really useful to recode the levels of one or several columns! 

```{r}
library(dplyr)

test_data <- data.frame(area = rep(c("North", "Sud", "East", "West"),times = c(2,3,4,1)),
                        quality_before = c("High","Low",
                                    "High","Low","Medium",
                                    "Medium","Low","High","High",
                                    "Low"),
                        quality_after = c("High","High",
                                    "High","Medium","Medium",
                                    "Low","Low","High","High",
                                    "Low"))

test_data %>%
  mutate(across(.cols = c(quality_before, quality_after),
            ~ case_when(
              . == "Low" ~ 0,
              . == "Medium" ~ 1,
              . == "High" ~ 2
            )
          )
  )
         
```


Strongly suggest to have a look at other functions and applications to perform column-wise operations https://cran.r-project.org/web/packages/dplyr/vignettes/colwise.html.


# Fitting several linear models by group with `purrr` 

For the example I am going to use the `flights` dataset from the R package `nycflights13`. I am going to fit linear model that tries to explain the `arr_time` as a function of `dep_time` and `arr_delay`.

```{r}
library(nycflights13)
library(purrr)

flights %>%
  dplyr::select(arr_time, dep_time, arr_delay, carrier) %>%
  head()
```


To fit the model to the whole dataset we would use the following code:

```{r}
summary(lm(arr_time ~ dep_time + arr_delay, data = flights))
```

What if we wanted to fit separate models by `carrier`?

```{r}
models <- flights %>%
  dplyr::select(arr_time, dep_time, arr_delay, carrier) %>%
  tidyr::nest(-carrier) %>%
  dplyr::mutate(fit = purrr::map(data, ~ lm(arr_time ~ dep_time + arr_delay, data = .))) %>%
  dplyr::mutate(results_fit = purrr::map(fit, function(f) confint(f))) 
models
```

```{r}
expand_models <- models %>%
  tidyr::unnest(results_fit, .drop=TRUE) 
expand_models
```

```{r fit-column}
expand_models$fit[1]
```

