---
title: "Basic but Useful R Tips by Song"
author: "Yutong Yuri Song"
date: "12/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Before we start

```{r}
sessionInfo()
```


## Data Exploration

I code in **tidyverse** universe plus **tidylog** to output all message corresponding to changes made to vector, dataframe, tibble, etc. Please find tidylog package @ <https://github.com/elbersb/tidylog>.

- An initial step is often to preview a dataset. An alternative to summary is describe in **Hmisc** package. The benefits will be it counts number of NAs as summary does, but it also show a frequency table for factor/character.


```{r, message=FALSE}
library(tidyverse)

# colnames(iris)
# summary(iris)
# str(iris)

library(Hmisc)

iris$Species %>% describe
```

- **tidylog** reduces significant code verification and avoid many errors for me. It warns you when NA is generated due to situation not considered. 

```{r}
library(tidylog, warn.conflicts = FALSE, quietly = FALSE)

new_dt <- iris %>% 
  filter(Sepal.Length >= 4.6) %>% 
  mutate(new_name = case_when(
    Species == "versicolor" ~ "V",
    Species == "setosa" ~ "S"))
```

- Compare with revised code: no new NA was generated after the mutation.


```{r}
library(tidylog, warn.conflicts = FALSE, quietly = FALSE)

new_dt <- iris %>% 
  filter(Sepal.Length >= 4.6) %>% 
  mutate(new_name = case_when(
    Species == "versicolor" ~ "Versicolor",
    Species == "setosa" ~ "Setosa",
    TRUE ~ "Virginica"))
```



## Data Preparation

Use iris dataframe as an example.

- **relocate** also support **.after**, and combine use with **where**.

```{r}
dt <- head(iris,5)

# dt %>% select("Species", everything(.))
dt %>% relocate("Species", .before = "Sepal.Length")
# dt %>% relocate(where(is.numeric), .after = where(is.factor))
```

- Similar concept can be applied to a vector through **SOfun** package. I found this useful when adjusting factor levels. Of course, **fct_reorder** and **fct_relevel** are useful in different situations.

```{r}
library(SOfun)
v <- letters[1:7]
v %>% moveMe(., "a last; b, e, g before d; c first; g after b")

```


- Another high frequency task is to manage NAs. This is way to spot hidden NAs.

```{r}
hidden_na_dt <- data.frame(
  "student" = rep(c("A", "B", "C"),2),
  "assignment" = rep(c("A1", "A2"),3),
  "mark" = c(NA, runif(n = 5, min = 45, max = 100))
) %>% 
  filter(!is.na(mark))

hidden_na_dt
```

- apply **complete** from **dplyr** package to fill 0 in missing mark from assignment 1 for student A. If there is more combinations, multiple items can be nesting in the **complete** argument.

```{r}
hidden_na_dt  %>% 
  complete(student, nesting(assignment), fill = list(mark = 0))
```

## Data Visualisation

I believe **ggplot2** / **plotly** is relative popular in practice. I also recommend **highercharter** to visualize timeseries data and/or **visNetwork** / **igraph** / **ggraph** to visualize networks. My focus today is labeling inside a chart, so that I will use **ggplot2** to demonstrate.

- Randomly picked a few countries by max number of population of that country to show potential difference when treating labeling.

```{r, message=FALSE}
plt_original <- population %>% 
  filter(country %in% c("India", "United States of America", "Viet Nam",
                        "Lao People's Democratic Republic")) %>% 
  ggplot(aes(x = year, y = population, group = country, color = country))+
  geom_line()

plt_original
```


### Functions that I used to improve numeric formatting

The purpose of having customized functions is to improve readability and reduce cognitive load for digesting information provided by visualization.

- The function I grabbed from stackoverflow and made two adaptations: (1) allow the function to accept negative inputs and (2) expand to recognize trillions. 

```{r, message=FALSE, warning=FALSE}
si_num <- function (x) {
  
  if (!is.na(x)) {
    
    if (x < 0){ 
      sign <-  "-"
      x <- abs(x)
    }else{
      sign <-  ""
      x <- x
    }
    
    if (x >= 1e12) { 
      chrs <- strsplit(format(x, scientific=12), split="")[[1]];
      len <- chrs[seq(1,length(chrs)-12)] %>% length();
      rem <- chrs[seq(1,length(chrs)-11)];
      rem <- append(rem, ".", after = len) %>% append("T");
    }
        
    if (x >= 1e9) { 
      chrs <- strsplit(format(x, scientific=12), split="")[[1]];
      len <- chrs[seq(1,length(chrs)-9)] %>% length();
      rem <- chrs[seq(1,length(chrs)-8)];
      rem <- append(rem, ".", after = len) %>% append("B");
    }
    
    
    else if (x >= 1e6) { 
      chrs <- strsplit(format(x, scientific=12), split="")[[1]];
      len <- chrs[seq(1,length(chrs)-6)] %>% length();
      rem <- chrs[seq(1,length(chrs)-5)]
      rem <- append(rem, ".", after = len) %>% append("M");
    }
    
    else if (x >= 1e3) { 
      chrs <- strsplit(format(x, scientific=12), split="")[[1]];
      len <- chrs[seq(1,length(chrs)-3)] %>% length();
      rem <- chrs[seq(1,length(chrs)-2)];
      rem <- append(rem, ".", after = len) %>% append("K");
    }
    
    else {
      return(x);
    }
    
    return(str_c(sign, paste(rem, sep="", collapse=""), sep = ""));
  }
  else return(NA);
} 

si_vec <- function(x) {
  sapply(x, FUN=si_num);
}
```


- Modifications include: (1) change graph title and axis titles and format, (2) change a theme: minimalist design,  (3) remove legend and add text labels to each line.
- Of course, there are more things: change color pallet defined for country, graph size,...

```{r, fig.height=5, fig.width=10,  message=FALSE, warning=FALSE}
library(hrbrthemes)
library(scales)
library(ggrepel)
library(cowplot)


year_series <- unique(population$year)
reminder <- (max(year_series) - min(year_series)) %% 4
new_breaks <- seq(from = min(year_series) + reminder, to = max(year_series), by = 4) 

df <- population %>% 
  filter(country %in% c("India", "United States of America", "Viet Nam",
                        "Lao People's Democratic Republic")) 
df_end <- df %>% 
  group_by(country) %>% 
  filter(year == max(year)) %>% 
  ungroup()

plt_adjust <- df %>% 
  ggplot(aes(x = year, y = population, group = country, color = country))+
  geom_line()+
  geom_point()+
  geom_text_repel(
    data = df_end,
    aes(label = str_wrap(country,25)),
    nudge_x = 1,
    direction = "y",## nudge vertically
    size = 3,
    hjust = 0, ### left aligned
    segment.size = 0.3, ### from here is about the line to connect the data point and text
    min.segment.length = 0,
    segment.color = "grey60") + 
  theme_ipsum() +
  theme(legend.position = "none") +
  scale_y_continuous(labels = si_vec)+
  scale_x_continuous(breaks = new_breaks, limits = c(NA, 2020))+
  labs(x = "Year", y = "Population", title = "Population Growth between 1995 and 2013")


plt_original
plt_adjust
```

- Or, put it into **plotly**, the default hover over message often does not satisfy users, more professional format is recommended to be used in hover over text.

```{r, message=FALSE, warning=FALSE}
library(plotly)

plt_plotly <- df %>% 
  mutate(text = str_c("Country: ", country, "\n",
                      "Year: ", year, "\n",
                      "Population: ", si_vec(population))) %>% 
  ggplot(aes(x = year, y = population, group = country, color = country, text = text))+
  geom_line()+
  geom_point()+
  theme_ipsum() +
  theme(legend.position = "none") +
  scale_y_continuous(labels = si_vec)+
  scale_x_continuous(breaks = new_breaks)+
  labs(x = "Year", y = "Population", title = "Population Growth between 1995 and 2013")



ggplotly({plt_plotly}, tooltip = "text")
```

### At the end

To be continue, I have coded many interactive plots in shinyapps, and some can be found from <https://coffeeandplot.com/apps/>. This is a relatively new website we created couples of month ago. Get in touch if you have any suggestions. Please find me @ <https://www.linkedin.com/in/ytsong/>.
